### 2018-03-17 paper/3-abstractions.tex
Add example for track

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  single linear chain. It is interesting to note that, when run on a
  \hs{Task}~\hs{Functor}, the \hs{busy} build system will build each key at most
  once, thus partially fulfilling the minimality requirement~\ref{def-minimal}.
  Alas, it still has no mechanism to decide which input keys changed since the
  previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\end{itemize}

@@

\vspace{1mm}
\definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
build task description, \hs{key} is a target key, \hs{store} is an initial
store, and \hs{result} is the store produced by running the build system with
parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
our abstractions:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, @@result :: Store i k v
result = @@build @@task @@key @@store
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
    \item \hs{store} and \hs{result} \emph{agree on inputs}, that is, for all
    input keys \hs{k}~$\in$~$I$:
    \[
    \hs{getValue}~\hs{k}~\hs{store}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}.
    \]
    In other words, no inputs were corrupted during the
    build\footnote{\todo{SLPJ}{Add the joke that GHC used to delete inputs?}}.

    \item The \hs{result} is \emph{consistent} with the \hs{task}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{Just}~\hs{(}\hs{getValue}~\hs{result}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{3mm}

@@

The above definition of correctness needs to be adjusted to accommodate
non-deterministic tasks and shallow cloud builds; we address this
in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects}, respectively.

\subsection{Computing dependencies}\label{sec-deps}

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task key = case task (\k -> Const [k]) key of
                            Nothing         -> []
                            Just (Const ks) -> ks
\end{minted}
\vspace{1mm}
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}.  So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

@@

\noindent
The implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record addition information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
\hs{runWriterT} to the contents of \hs{Maybe}.

% Here is an example of \hs{track}ing monadic tasks:

% \vspace{1mm}
% \begin{minted}[xleftmargin=10pt]{haskell}
% @\ghci@ store1 = initialise () (\key -> if key == "A1" then 10 else 1)
% @\ghci@ track sprsh2 store1 "B1"
% Just (1,["C1","B2"])
% \end{minted}
% \vspace{1mm}
% \begin{minted}[xleftmargin=10pt]{haskell}
% @\ghci@ store2 = initialise () (\key -> if key == "A1" then 10 else 20)
% @\ghci@ track sprsh2 store2 "B2"
% Just (20,["C1","B1"])
% \end{minted}
% \vspace{1mm}

% \noindent
% As expected, the dependencies of \hs{sprsh2} (see the spreadsheet
% in~\S\ref{sec-task-monad}) are determined by the contents of the store.
% Furthermore, the above example highlights the static cycle between cells
% \cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
% whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.

>>>>

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  single linear chain. It is interesting to note that, when run on a
  \hs{Task}~\hs{Functor}, the \hs{busy} build system will build each key at most
  once, thus partially fulfilling the minimality requirement~\ref{def-minimal}.
  Alas, it still has no mechanism to decide which input keys changed since the
  previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\end{itemize}

@@

\vspace{-1mm}
\definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
build task description, \hs{key} is a target key, \hs{store} is an initial
store, and \hs{result} is the store produced by running the build system with
parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
our abstractions:

\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, @@result :: Store i k v
result = @@build @@task @@key @@store
\end{minted}

@@

\begin{itemize}
    \item \hs{store} and \hs{result} \emph{agree on inputs}, that is, for all
    input keys \hs{k}~$\in$~$I$:
    \[
    \hs{getValue}~\hs{k}~\hs{store}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}.
    \]
    In other words, no inputs were corrupted during the
    build.
    %\footnote{\todo{SLPJ}{Add the joke that GHC used to delete inputs?}}.

    \item The \hs{result} is \emph{consistent} with the \hs{task}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{Just}~\hs{(}\hs{getValue}~\hs{result}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{2mm}

@@

% The above definition of correctness needs to be adjusted to accommodate
% non-deterministic tasks and shallow cloud builds; we address this
% in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects}, respectively.

\subsection{Computing dependencies}\label{sec-deps}\label{secdeps}

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task key = case task (\k -> Const [k]) key of
                            Nothing         -> []
                            Just (Const ks) -> ks
\end{minted}

Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}.  So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

@@

\noindent
The implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
\hs{runWriterT} to the contents of \hs{Maybe}.
Here is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
C1: 1
B2: 10
(10,["C1","B2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
C1: 2
A2: 20
(20,["C1","A2"])
\end{minted}
\vspace{1mm}

\noindent
As expected, the dependencies of the cell \cmd{B1} from \hs{sprsh2} (see the
spreadsheet in~\S\ref{sec-task-monad}) are determined by the value of \cmd{C1},
which in this case is obtained by reading from the standard input.

% Furthermore, the above example highlights the static cycle between cells
% \cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
% whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.
